---
title: "Ex. Assignment Solutions"
author: "Andrew"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, broom)
```

# [Ex:]{.ex} Homework question

We are going to learn how to use R Markdown to write our homework [and]{.hi} by doing a fake homework question together. This will also serve as instruction for three important (base) functions for running regressions. The three functions are:

- __`lm`__: 
  - *Description**: Stands for "linear model." It is used to fit linear models to data. The function calculates the coefficients of a linear equation, involving one or multiple predictors that best predict the outcome (dependent variable). 
  - [Ex.]{.ex} `model = lm(y ~ x, data = data)`

- __`resid`__: 
  - *Description*: Extracts the residuals from a model. Residuals represent the difference between the observed value and the value predicted by the model.
  - [Ex.]{.ex} `model_residuals = resid(model)`

- __`predict`__: 
  - *Description*: Given a model and new data, this function produces predicted values based on the model's coefficients. It's often used to see how well a model performs on new, unseen data or to predict outcomes for specific input values.
  - [Ex.]{.ex} `predicted_values = predict(model, newdata = new_data)`


Understanding these three functions will serve as a solid foundation for conducting regression analyses in R.

Before we begin, we'll need to generate the data with the following

```{r}
# Set random seed to ensure results are consistent
set.seed(42)

# Generating data
  # Number of observations
  n = 200  
  # Years of education as random variable, mean 12 years, standard deviation 2 years
  years_of_education = rnorm(n, mean=12, sd=2)
  # Income as a random variable
  income = 5000 + 2000*years_of_education + rnorm(n, mean=0, sd=500 + 250*years_of_education)

# Create a tibble
tbl = tibble(years_of_education, income)
```

[Q1.]{.question} **Data Exploration:**

- Create a scatter plot with `years_of_education` on the x-axis and `income` on the y-axis. What does the relationship between the two variables look like?

```{r}
plot(tbl, xlab="Years of Education", ylab="Income", 
     main="Income vs Years of Education")
```


[Q2.]{.question} **Simple Linear Regression:**

- Write the proper formula that describes the standard wage regression.

$$
\text{Income}_i = \alpha + \beta \; \text{YearsEdu}_i + \varepsilon_i
$$

- Using the `lm` function, run a regression of `income` on `years_of_education`. Assign your linear model to a variable called `model`.  

```{r}
model = lm(income ~ years_of_education, data = tbl)
```

[Q3.]{.question} **Model Summary:**

- Extract and interpret the coefficients of the regression using the `broom` package function `tidy`

```{r}
tidy(model)
```

- What does the coefficient of `years_of_education` tell you about the relationship between education and income?

For each additional year of education, there is an associated increase in income by \$1865

- Are the coefficients statistically significant at the 5% level? Provide evidence for your answer.

Yes, the we can reject the null hypothesis that year of education has no correlation with income since the p value is $3.65e-34$ which is $< 0.05$

[Q4.]{.question} **Residuals:**

- Suppose we are worried about heteroskedasticity. What is a simple, not so scientific way to check for heteroskedastic errors?

A quick and dirty check is to plot the residuals across $x$ (or in this case years of education)

- Plot the residuals and discuss

```{r}
model_resid = resid(model)

plot(tbl$years_of_education, model_resid, 
     xlab="Years of Education", ylab="Residuals", 
     main="Residuals vs Years of Education")
```


[Q5.]{.question} **Predictions:**

- Use the `predict` function to estimate the income of an individual with 10, 14, and 18 years of education.   
- How do these predictions compare with the original simulated relationship?  

```{r}
new_tbl = tibble(years_of_education = c(10, 14, 18))
predicted_income = predict(model, newdata = new_tbl)

new_tbl %>% mutate(prediction = predicted_income)


```

